In the Age of Web:\\ \textnormal{\LARGE Typed Functional-First Programming Revisited}

### 2015-01-06 ml-2014-paper/paper.tex
Complete first version

\usepackage{graphicx}
\usepackage{xcolor}
\usepackage[fleqn]{amsmath}
\usepackage[hang,flushmargin]{footmisc} 
\usepackage{graphics} 
%\usepackage{breakurl}

@@

\newcommand{\ignp}{\_\kern-.5ex\_\kern-.5ex\_ }
\newcommand{\kvd}[1]{\textnormal{\textcolor{kvdclr}{\sffamily #1}}}
\newcommand{\str}[1]{\textnormal{\textcolor{strclr}{\ttfamily "#1"}}}
\newcommand{\ident}[1]{\textnormal{\sffamily #1}}
\newcommand{\lident}[1]{\textnormal{\sffamily 
  \`{}\hspace{-0.25em}\`{}\hspace{-0.1em}#1\`{}\hspace{-0.25em}\`{}}}
\newcommand{\cmt}[1]{\textit{\sffamily\textcolor{cmtclr}{#1}}}
\newcommand{\narrow}[1]{\hspace{-0.75em} #1 \hspace{-1.25em}~}

@@

% What is my solution
In this paper, we present how F\# language and libraries face the challenges posed by the web.
Technically, this comprises using \emph{type providers} for integration with external information 
sources and for integration with untyped programming environments, using \emph{lightweight 
meta-programming} for targetting JavaScript and \emph{computation expressions} for writing 
asynchronous code.

@@

@@

This additional flexibility makes it possible to use F\# in ways that break the common assumptions
that are often taken for granted in languages such as ML and Haskell\footnote{In a way, we are trying
to uncover the hidden assumptions of the functional programming \emph{research programme} that are
normally ``\emph{rendered unfalsifiable by the methodological decisions of its protagonists}.'' 
(Lakatos \cite{philosophy-lakatos} as quoted by Chalmers \cite{philosophy-thing})}. The focus on 
the web directs our inquiry and provides an angle for reconsidering such assumptions.

@@

@@

\subsection{Loading data and updating the user interface}

@@

The aim of this paper is not to make value judgements and argue what is better. Using the case
study as a basis, we claim that the outlined approach is \emph{one possible} and that it
\emph{works in practice}. In the rest of the paper, we give more details about the most important 
aspects of the approach and discuss alternatives`'.

>>>>

\usepackage{graphicx}
\usepackage{xcolor}
\usepackage[fleqn]{amsmath}
\usepackage[hang,flushmargin]{footmisc} 
\usepackage{graphics} 
\usepackage{stmaryrd}
\usepackage{amsthm}
%\usepackage{breakurl}

@@

\newcommand{\sem}[1]{\llbracket #1 \rrbracket}
\newcommand{\ignp}{\_\kern-.5ex\_\kern-.5ex\_ }
\newcommand{\kvd}[1]{\textnormal{\textcolor{kvdclr}{\sffamily #1}}}
\newcommand{\str}[1]{\textnormal{\textcolor{strclr}{\ttfamily "#1"}}}
\newcommand{\ident}[1]{\textnormal{\sffamily #1}}
\newcommand{\lident}[1]{\textnormal{\sffamily 
  \`{}\hspace{-0.25em}\`{}\hspace{-0.1em}#1\`{}\hspace{-0.25em}\`{}}}
\newcommand{\cmt}[1]{\textit{\sffamily\textcolor{cmtclr}{#1}}}
\newcommand{\narrow}[1]{\hspace{-0.75em} #1 \hspace{-1.25em}~}

\newtheorem*{theorem*}{Theorem}

@@

% What is my solution
In this paper, we present how F\# language and libraries face the challenges posed by the web.
Technically, this comprises using \emph{type providers} for integration with external information 
sources and for integration with untyped programming environments, using \emph{lightweight 
meta-programming} for targeting JavaScript and \emph{computation expressions} for writing 
asynchronous code.

@@

% =================================================================================================
%
%                         #                #           
%   #       #               #            #               
%   # # #  #### # #  ##   ### #  #  ### #### #  ##  # #  
%   # ## #  #   ### #  # #  # #  # #     #   # #  # ## # 
%   # #  #  #   #   #  # #  # #  # #     #   # #  # #  # 
%   # #  #  #   #   #  # #  # #  # #     #   # #  # #  # 
%   # #  #  #   #   #  # # ## # ## #     #   # #  # #  # 
%   # #  #   ## #    ##   # #  # #  ###   ## #  ##  #  # 
%                                                    
% =================================================================================================

@@

This additional flexibility makes it possible to use F\# in ways that break the common assumptions
that are often taken for granted in languages such as ML and Haskell\footnote{In a way, we are trying
to uncover the hidden assumptions of the functional programming \emph{research programme} that are
normally ``\emph{rendered unfalsifiable by the methodological decisions of its protagonists}.'' 
(Lakatos \cite{philosophy-lakatos}, quoted by Chalmers \cite{philosophy-thing})}. The focus on 
the web directs our inquiry and provides an angle for reconsidering such assumptions.

@@

% =================================================================================================
%                                                  
%                                            #      
%     ###                   ###  #           #      
%    #     ##    ##  ##    #    #### #  #  ### #  # 
%   #        #  #   #  #   #     #   #  # #  # #  # 
%   #      ###  ##  ####    ##   #   #  # #  # #  # 
%   #     #  #   ## #         #  #   #  # #  # #  # 
%    #    #  #    # #         #  #   # ## # ##  ##  
%     ###  ## # ##   ###   ###    ##  # #  # #  ##  
%                                               #   
%                                              ##   
% =================================================================================================

@@

\subsection{Loading data and updating the user interface}
\label{sec:case-loading}

@@

The aim of this paper is not to make value judgements and argue what is better. Using the case
study as a basis, we claim that the outlined approach is \emph{one possible} and that it
\emph{works in practice}. In the rest of the paper, we give more details about the most important 
aspects of the approach and discuss alternatives.

% =================================================================================================
%
%                                            #    # 
%     ##                    #     #          #    # 
%    #  #  # #   ##  # #    #  #  #  ##  # # #  ### 
%   #    # ## # #  # ## #   #  #  # #  # ### # #  # 
%   #    # #  # #### #  #   # # # # #  # #   # #  # 
%   #    # #  # #    #  #    #   #  #  # #   # #  # 
%    #  #  #  # #    #  #    #   #  #  # #   # # ## 
%     ##   ###   ### #  #    #   #   ##  #   #  # # 
%          #                                        
%          #                                        
% =================================================================================================

\section{Integrating with the open world}
\label{sec:tp}

Type providers \cite{fsharp-typeprov} is a mechanism for integrating external components into a 
statically typed programming language. Such components include information sources (such as World
Bank), other environments (here, JavaScript libraries via TypeScript), but type-providers can 
also be used for limited meta-programming (the \ident{Asynchronous} parameter can be seen as
a form of meta-programming).

\subsection{How type providers work and fail}
\label{sec:tp-def}

In technical terms, type providers are libraries that are loaded by the compiler (and editor)
and are executed at compile-time (or edit-time). A type provider builds information about types
it provides and makes those available to the compiler. It is worth noting that this is done lazily,
so the type provider does not need to provide types for the entire information space at once.

In terms of programming language theory, type providers can be easily explained anecdotally. 
They change the starting point for a type-checking of a program as follows:
%
\begin{equation*}
\hspace{8em}
\begin{array}{rclll}
 \emptyset &\narrow{\vdash}& e : \tau 
    &\qquad& (\textit{classical, or closed-world})\\
 \pi~(\includegraphics[width=1em,trim=0 1.5em 0 0]{earth.png})  &\narrow{\vdash}& e : \tau
    &\qquad& (\textit{type proiders, or open-world})
\end{array}
\end{equation*}
%
Rather than starting the type-checking with an empty context, we start type-checking with a
context containing a \emph{projection} of some information from the world. In this paper, we do not
attempt to formalize type providers, so we avoid the question of what, formally, is the world here.
However, this analogy is useful for understanding how type providers work and how they can go wrong:

\begin{itemize}
\item \emph{Type provider failure.} 
  The projection (a type provider) is implemented in F\# and can fail (for example, if internet
  is required but unavailable). However, when it succeeds, it generates a valid F\# code and so 
  unchecked compilation errors cannot happen.

\item \emph{Runtime failure.}
  When the world changes and we \emph{run} the program, we may get a runtime exception. It is
  expected that the provided code will not fail as long as certain assumptions are satisfied (for
  example, countries are not removed from a database). But if the world changes and the assumptions
  no longer hold, the provided code may fail with a runtime exception.

\item \emph{Recompilation failure.} 
  When the world changes and we \emph{recompile} the program, the result of the projection can
  differ and so we may not be able to type-check code that used to type-check. This can be seen as
  negative aspect, but there is a clear benefit -- it means we discover an error that would happen
  at runtime earlier, that is during recompilation.
\end{itemize}

\noindent
The above points are \emph{general} points about type providers. In this paper, we are more interested
in the \emph{specific} type providers used in the case study. We look at the World Bank and TypeScript
type providers in the following two sections. When discussing type providers, we need to look at 
how the provider (or the projection $\pi$) works and what assumptions it makes about the world. 
That is, what world changes cause the provided type to change and what world changes cause a 
runtime exception.

\subsection{Integrating with World Bank data}
\label{sec:tp-data}

The World Bank provider \cite{fsharp-data} is a type provider for a specific data source -- it has 
been designed specifically for World Bank data. This contrasts with F\# Data type providers for 
working with CSV, XML and JSON data that take sample or schema as parameter and can thus be used 
with files of any structure.

\subsubsection{The World Bank type provider}

The projection implemented by the type provider generates a type called \ident{Countries} that contains 
countries (and regions) as members. Each member returns a value of type \ident{Indicators} that is
also generated and contains all indicators as members. Note that the type provider generates just 
these two types -- in particular, it does not generate a new type for each country. This would be an 
interesting option if we could check what indicators are available for a given country.

The type provider builds the structure of the provided types, but the provided types are \emph{erased} 
during compilation and replaced with runtime implementation. For the World Bank type provider, the 
erasure works as follows:
%
\begin{equation*}
\begin{array}{lcl}
 \sem{~\ident{data.}\lident{Czech Republic}~} &=&
   \ident{data.GetCountries}().\ident{GetCountry}(\str{CZE}) \\
 \sem{~\ident{cz.}\lident{School enrollment, tertiary (\% gross)}~} &=&
   \ident{cz.AsyncGetIndicator}(\str{SE.TER.ENRR})
\end{array}
\end{equation*}
%
The underlying operations (\ident{GetCountry}, \ident{GetCountries} and \ident{AsyncGetIndicator})
are operations of an underlying runtime library (which is a normal non-provided library). This is
a typical pattern -- type providers generally produce a light layer on top of a rich runtime library.

An important point here is that the type provider uses the \emph{name} of a country or an indicator
for the member name, but at runtime, it uses a \emph{code} of the country or the indicator. This has
interesting implications for the safety properties, as discussed in the next section.

\subsubsection{Safety properties of the type provider}

Now that we know how the World Bank type provider generates the types and the runtime code, we can 
look at the assumptions it makes about the world and in which cases it encounters one of the failures
defined in Section~\ref{sec:tp-def}.

\begin{itemize}
\item If you are offline when using the type provider (compiling or editing), the type provider will
  not be able to obtain the list of countries or indicators. This is an example of \emph{type provider
  failure}. In case of World Bank, this is partly prevented by caching of the schema, but internet
  access is required when using the provider for the first time.

\item An interesting case is when a country or indicator is renamed. In that case, compiled code
  using the World Bank type provider will continue to work (there is no \emph{runtime failure}), 
  because the lookup uses the country or the indicator code. However, recompilation will fail
  as the member name (which appears in the source code) will be different.

\item Finally, if a country or indicator completely disappears, we will get \emph{runtime failure}
  when running existing compiled code and \emph{recompilation failure}. here, the \emph{recompilation
  failure} is a useful indicator that the assumptions made by our code have been violated by
  the data source.
\end{itemize}

\noindent
When contrasted with the traditional view of safety in ML family of languages, the World Bank 
type provider certainly relaxes the usual safety conditions. However, this is not the case when 
compared with standard developer practices. In particular, if we wrote 
$\ident{data.GetCountries}().\ident{GetCountry}(\str{CZE})$ by hand, the \emph{runtime failure}
behaviour would be the same, but we would lose the useful \emph{recompilation failure}. The following
section discusses this (and other) alternatives.

\subsubsection{Discussion of alternatives}

There are two main alternatives that are worth discussing. First, how would the code look if 
we did not use any type provider. Second, what are alternative designs for the World Bank type 
provider and how would such alternatives change the safety properties.

\vspace{-1em}
\paragraph{Accessing countries without type providers.} First, consider how would we write the code to obtain a list
of countries without using the World Bank type provider. Assuming we want only the countries and
regions in Figure~\ref{fig:wb}, we could specify the country codes and then obtain the country objects:
%
\begin{equation*}
\begin{array}{l}
 \kvd{let}~\ident{data}~=~\ident{WorldBank.GetDataContext}() \\
 \kvd{let}~\ident{countryCodes}~=~\lbrack~\str{EU},~\str{CZE},~\str{GBP},~\str{USA}~\rbrack \\
 \kvd{let}~\ident{countries}~=~\ident{countryCodes}~|\rangl~\ident{List.map}~(\ident{data.GetCountries}().\ident{GetCountry})
\end{array}
\end{equation*}
%
The motivation for the case study was a data journalism application with a transparent logic
that is easy to understand (and perhaps even modify). The type provider certainly achieves these
goals better -- a country can be added just by looking through an auto-complete list and we do not
need to know its code or exact name. However, if we wanted to list all countries in a region, then
the World Bank type provider would not be helpful.

As for the safety properties, as discussed earlier, using explicit country codes has the same
runtime properties as using the type provider. This could be avoided by listing all countries in
a given region (or, according to another rule), but that is solving a different problem.

In summary, the World Bank type provider fits a specific niche that we were exploring in the case 
study. That is, when we want to create an information analysis that accesses specific data from
an information source. With rising popularity of data journalism and Open Government Data
initiatives, we argue that this is an important problem domainHere, we first specify the country 
codes and then obtain the country objects.

\vspace{-1em}
\paragraph{Providing safer data accessors.} 
The \emph{recompilation failure} when a country is removed from the dataset is an indication that
code would not behave as expected as runtime and so we find it useful. The problematic case is 
a \emph{runtime failure} when a country is removed after the program is compiled. One way to 
avoid this error would be to change the type of the provided members like
\ident{data.}\lident{Czech Republic} from \ident{Country} to \ident{Country~option}.

However, that is only shifting the burden of error handling from the library to the user. If the
author of the application does not intend to handle errors (for example, by skipping missing data),
they would have to write the following:
%
\begin{equation*}
\begin{array}{l}
 \kvd{let}~\ident{optCountries}~= \\
 \quad~\lbrack~~ \ident{data.Countries.}\lident{European Union} \\
 \qquad   \ident{data.Countries.}\lident{Czech Republic} \\
 \qquad   \ident{data.Countries.}\lident{United Kingdom} \\
 \qquad   \ident{data.Countries.}\lident{United States} ~\rbrack 
\\[0.5em]
 \kvd{let}~\ident{countries}~=\\
  \quad \ident{optCountries}~|\rangl~\ident{List.map}~(\kvd{fun}~\ident{countryOpt}~\rightarrow \\
  \qquad \kvd{match}~\ident{countryOpt}~\kvd{with}~\ident{Some~c}~\rightarrow~\ident{c}~|~\ignp~\rightarrow 
    \ident{failwith}~\str{Country does not exist!})
\end{array}
\end{equation*}
%
For the purpose of our case study, providing option types would only make the code more complex.
However, if we wanted to handle errors more gracefully, then option types would be a better 
alternative. Note, however, that we can still handle exceptions using $\kvd{try}~\ldots~\kvd{with}$.

Whether using option types is more desirable or not often depends on the use case (how we
want to handle errors) and reliability of the data source (countries do not disappear often).
This alternative can be supported by adding a static parameter to the type provider (like
$\ident{Asynchronous}=\kvd{true}$). For example, the CSV type provider from F\# Data has a 
parameter \ident{AssumeMissingValues} that serves exactly this purpose.

\vspace{-1em}
\paragraph{Accessing data ``as of time''.} 
So far, we discussed how to mitigate the problems caused by the fact that the open world changes.
Can we make the world \emph{not} change? This is a sensible question for some data sources
(although not for the World Bank). For example Freebase \cite{google-freebase} supports the 
notion of ``as of time''. When calling Freebase with the \ident{as\_of\_time} parameter, the API
returns data and meta-data that was present at the specified date.

This would be a desirable option for our case study -- if we want to illustrate point made 
by an article, the accompanying application could use data available at the time when the 
article was written. However, this feature relies on the ability of the data source (not all
services have this), or on the possibility to create local snapshots (difficult for large data
sources).

\subsection{Integrating with JavaScript libraries}
\label{sec:tp-lang}

Type providers are often described as a technique that simplifies data access. The TypeScript
type provider shows that there is a broader range of uses, including integration between different
programming languages. Another example of type provider for language integration is the R
provider \cite{fsharp-rprovider} which imports packages and functions of the statistical 
environemnt R.

\subsubsection{The TypeScript type provider}
The projection implemented by the TypeScript type provider imports interface definitions in the
\str{.d.ts} format specified by TypeScript \cite{ms-typescript}. This is a format of type 
annotations for JavaScript libraries. The following example is an excerpt from the \str{jquery.d.ts}
file that describes the \ident{jQuery} value and the \ident{attr} method used in our example:
%
\begin{equation*}
\begin{array}{l}
 \kvd{declare var}~\ident{jQuery}~:~\ident{JQueryStatic}; 
\\[0.5em]
 \kvd{interface}~\ident{JQueryStatic}~\{ \\
 \quad (\ident{selector}: \kvd{string}, \ident{context?}: \kvd{any}):~\ident{JQuery}; \\
\} 
\\[0.5em]
\kvd{interface}~\ident{JQuery}~\{ \\
\quad \ident{attr}(\ident{attributeName}~:~\kvd{string}):~\kvd{string}; \\
\quad \ident{attr}(\ident{attributeName}~:~\kvd{string},~\ident{value}~:~\kvd{any}):~\ident{JQuery}; \\
\}
\end{array}
\end{equation*}
%
The example demonstrates a number of typical problems that arise when using type providers for
language interoperation. Not all TypeScript language constructs have direct equivalent in F\# and 
the TypeScript type provider needs to map them to other constructs that are available:
%
\begin{itemize}
\item The interface file defines \ident{jQuery} as a global variable. Although F\# supports
  global bindings, those cannot be provided by a type provider and so the global variable
  is exposed as a static member of the imported type, so we use it via \ident{j.jQuery}.

\item The interface \ident{JQueryStatic} specifies that the \ident{jQuery} object is callable
  (other members are omitted). This denotes that the corresponding JavaScript object is a
  callable function with other members. F\# does not allow ``calling an object'' and so 
  this is mapped to a method \ident{j.jQuery.Invoke}. 
\end{itemize}

\noindent
It is also worth noting that the \ident{attr} member is overloaded and uses an optional parameter.
However, both of these features are available in F\# and can be directly mapped. Now that we 
covered how the type structure is generated, we also need to show what code is generated when 
the types are erased:
%
\begin{equation*}
\begin{array}{l}
 \sem{~\ident{j.jQuery.Invoke}(\ident{command})~} ~= \\
 \qquad \kvd{let}~\ident{jq}~=~\ident{Emit.PropertyGetImpl}(\kvd{true}, \str{jQuery}, \lbrack|~|\rbrack) \\
 \qquad \ident{Emit.CallImpl}(\kvd{false}, \str{}, \lbrack|~ \ident{jq};~ \ident{command}~|\rbrack~) \\[0.75em]
 \sem{~\ident{jQuery}(\str{<input>})\ident{.attr}(\str{type},\str{checkbox})~} ~= \\
 \qquad \ident{Emit.CallImpl}(\kvd{false}, \str{attr}, \lbrack|~ \ident{jQuery}(\str{<input>});~\str{type};~\str{checkbox}~|\rbrack~)
\end{array}
\end{equation*}
%
An important point here is that the above provided code is never actually \emph{executed}. 
As discussed in Section~\ref{sec:js-meta}, it is translated to JavaScript (which then runs
in the web browser). So, the above code can be seen more as instructions for the translator.
The mapping is quite simple -- member calls are translated to \ident{Emit.CallImpl} (empty
name denotes that the object itself is called) and property getters are translated to 
\ident{Emit.PropertyGetImpl}. In both cases the first argument denotes whether the call is
static and the last argument is an array of arguments.

\subsubsection{Safety of cross-language type providers}
As mentioned earlier, TypeScript does not have the type soundness property \cite{ms-safets},
because of covariant generics. This illustrates two points. First, this is a property of
\emph{running} TypeScript code, which is not the case here -- we use TypeScript only as a
source of annotations for JavaScript libraries. There is no guarantee that the library will 
actually adhere to the specification, which is an issue that we return to when discussing
translation to JavaScript (Section~\ref{sec:js-meta}).

The second point is that we can not import the typing rules associated with TypeScript.
The type provider mechanism can only provide F\# types that then behave according to the
F\# typing rules. This means that we can import generics\footnote{Due to technical 
limitations, this is not currently allowed by F\# type providers, but there are no
fundamental reasons for this.}, but this will not automatically allow using them 
covariantly. In some cases, the F\# type system is simply more strict -- and in that case,
users have to use unsafe operations, such as $\ident{unbox}\langl \ident{bool} \rangl$ 
in the \ident{render} function. We return to the topic of unsafe operations in 
Section~\ref{sec:js-meta}.

The above discussion highlights a broader point about using type providers for language
interoperability. Given any language, its type system be weaker or stricter in some ways.
Both cases make mapping difficult:

\begin{itemize}
\item First, consider a weaker system with less type information. The type provider may need
  to map more types to \ident{object} (or similar general type). However, this will make 
  the provided operations hard to use (because the source language is more flexible than
  F\#). An alternative is to require explicit annotations (like \textcolor{strclr}{\ttfamily d.ts} 
  files for JavaScript).

\item Second, if the imported language has a more precise type system than F\#, the type 
  provider hsa to drop some of the information. This can be done safely for types in 
  contravariant positions. In the unsafe case, the runtime needs to perform dynamic checks 
  (Section~\ref{sec:conc-interop}).
\end{itemize}

\noindent
In case of JavaScript, we encounter both cases. In many cases, the return type is not
statically known and is exposed as \kvd{any}, which we then map to F\# \ident{object}
(and the developer has to use \ident{unbox}). However, TypeScript also supports limited
form of dependent typing (overloading on constants \cite{ms-typescript-09}). This cannot
be expressed in the F\# type system and the developer has to do more work (for example,
choose the right overload or provide an explicit type annotation).

An interesting example of type provider for a language with a weaker type system is the
R provider \cite{fsharp-rprovider}. It solves the mismatch by using runtime reflection 
to discover R functions and their declared parameters. However, R provider also generates
additional function overloads that let the user specify additional parameters (not explicitly 
declared in the signature). To make such interoperability easier, we propsed additional 
escape mechanism \cite{fsharp-paramsdict} that provide usable syntax when static type information 
cannot be fully recovered.

\subsubsection{Discussion of alternatives}
There are two options for calling JavaScript libraries such as jQuery. We can use annotations
-- written either in F\# or imported into F\#, or we can let the user write inline JavaScript. 
A different approach is to discourage the use of existing JavaScript libraries as discussed in 
Section~\ref{sec:js-lib}.

\vspace{-1em}
\paragraph{Embedding inline JavaScript.}
To make the interoperability mechanism general, the host language (here, F\#) should not know about 
the syntax of the embedded language (here, JavaScript). This means that the JavaScript code can
only be embedded as strings. FunScript provides this option and it is used for accessing various
primitive JavaScript operations. For example, the following is one way to define a function that
converts a value to a number using JavaScript semantics:
%
\begin{equation*}
\begin{array}{l}
 \lbrack\langl\ident{JSEmit}(\str{return \{0\}*1.0;})\rangl\rbrack \\
 \kvd{let}~\ident{number}~(\ignp:\ident{obj}) : \ident{float} ~=~\ident{failwith}~\str{JavaScript stub should not be called.}
\end{array}
\end{equation*}
%
The example uses .NET attributes (on the first line), which are meta-data attached to a function.
When FunScript finds a call to such annotated function, it replaces the call with the specified 
JavaScript. The string \textcolor{strclr}{\ttfamily \{0\}} is a placeholder for the argument. Note
that the body of the function is never actually executed and so we use a placeholder.

The approach provides no guarantees about the inline JavaScript code. At compile-time, it cannot
even check that the JavaScript code is syntactically correct. This could be improved using a 
parameterized type provider such as $\ident{JSCode}\langl\str{return \{0\}*1.0;}\rangl$, which 
would check that the static parameter is syntactically correct JavaScript code. So, type providers
could be used to make this approach safer.

\vspace{-1em}
\paragraph{Other ways of writing annotations.}
Many compilers to JavaScript provide a way for writing type signatures for JavaScript
libraries in the host language. For example, the following snippet shows a declaration of the
jQuery \ident{attr} method from a mapping for js\_of\_ocaml \cite{js_ocmal} (definitions in 
SMLtoJs \cite{js_sml} are similar, but simpler):
%
\begin{equation*}
\begin{array}{l}
 \kvd{class type}~\ident{jQuery}~=~\kvd{object} \\
 \quad \kvd{method}~\ident{attr}~:~\ident{js\_string t} ~\rightarrow~ \ident{js\_string t optdef meth} \\
 \quad \kvd{method}~\ident{attr\_set}~:~\ident{js\_string t} ~\rightarrow~ \ident{js\_string t} ~\rightarrow~ \ident{jQuery t meth} \\
 \kvd{end}
\end{array}
\end{equation*}
%
The js\_of\_ocaml project uses the OCaml object model for calling JavaScript libraries. 
Compared with FunScript, it is more explicit in importing JavaScript -- the \ident{meth}
type denotes a JavaScript method; \ident{optdef} specifies that the result may be 
undefined and \ident{js\_string} denotes a JavaScript string (which is distinct from OCaml
strings). Also, note that the second method is called \ident{attr\_set}. This is a simple
naming trick -- OCaml does not support overloading and js\_of\_ocaml simply ignores anything
after underscore.

Using a type provider is similar to writing a code generator that turns TypeScript 
\textcolor{strclr}{\ttfamily d.ts} files into the above OCaml annotation. The main difference
is that type providers provide a smoother integration. Type providers do not produce any artefacts
(generated files) and can import entire repositories lazilly.

Type providers for information sources, such as the World Bank provider, weaken the usual notion 
of type safety. In general, they require that certain assumptions about the world do not change
between the state of the world at compile-time and state of the world at run-time. The TypeScript
provider is different and it is more interesting to consider runtime behaviour as discussed
in Section~\ref{sec:conc-interop}.

For type providers that provide access to external information sources, we can formulate a 
\emph{relativized} notion of type safety. A full formalization is beyond the scope of this
paper, but the following provides an outline of such theorem.

\begin{theorem*}[Relativized type safety]
Assume $\pi_{\ident{~WorldBank}}$ is the mapping implemented by the World Bank type provider,
$w, w_0$ are models of a (read-only) world that can be seen as functions defined on country-indicator
pairs and $\langl e, w \rangl \rightarrow e'$ is a one-step reduction on expressions that has 
access to the World Bank data modelled by $w$. Then we define:

\begin{itemize}
\item
Multi-step reduction $\langl e, w \rangl \rightarrow^* e'$ if either $e=e'$
or $\langl e, w \rangl \rightarrow e''$ and $\langl e'', w \rangl \rightarrow^* e'$. 
\item
Functions $\ident{countries}(e)$ and $\ident{indicators}(e)$ that return sets of 
country names \\and indicator names that appear in an expression $e$. 
\end{itemize}
%
Then the following \emph{relativized type safety} implication holds:
%
\begin{equation*}
\begin{array}{lcl}
(~~\pi_{\ident{~WorldBank}}~(w_0) \vdash e : \tau 
  && \textnormal{If $e$ is well-typed using a compile-time world $w_0$}
\\
\quad \wedge~ \langl e, w \rangl \rightarrow^* e'
  && \textnormal{\ldots and the expression reduces in one or more steps}
\\
\quad \wedge~ (~\forall c \in \ident{countries}(e), 
  && \textnormal{\ldots and the run-time world model $w$ contains all}
\\
       \qquad~~~ \forall i \in \ident{indicators}(e) . \langl c, i \rangl \in \mathit{dom}(w)))
  && \textnormal{\;\quad country-indicator pairs that may be accessed,} 
\\
\Rightarrow ~\ident{value}(e') ~\vee~ (\exists e'' . \langl e', w \rangl \rightarrow e'') 
  && \textnormal{then the result is either a value or can further reduce.}
\end{array}
\end{equation*}
\end{theorem*}

\noindent
The theorem has the usual structure of a type safety theorem. The reduction is defined on pairs
consisting of expression and a world $w$. The result of the reduction is only an expression and
so it cannot modify the world. Even though the above is only a brief sketch, it demonstrates two
important points about relativized type safety that are relevant to many type providers.

First, we distinguish between the state of the world $w_0$ that is used at compile-time and the 
state $w$ used at run-time. If these were the same, the additional assumption would always hold.
Second, the additional assumptions states that all countries and indicators that \emph{may be
accessed} need to be available. This is stricter than necessary (akin to effect systems),
but it does not specify that the run-time world must contains \emph{all} provided countries 
and indicators.

% =================================================================================================
%
%                         # # #                                                            #           
%   ###                     #                #            #                   ###                  #   
%  #     ##  # #  #  # #  # # # # #   ###   ####  ##      #  ##   #  #  ##   #     ### # # # # #  #### 
% #     #  # ## ## # ## # # # # ## # #  #    #   #  #     #    #  #  #    #  #    #    ### # ## #  #   
% #     #  # #  #  # #  # # # # #  # #  #    #   #  #     #  ###  #  #  ###   ##  #    #   # #  #  #   
% #     #  # #  #  # #  # # # # #  # #  #    #   #  #     # #  #  #  # #  #     # #    #   # #  #  #   
%  #    #  # #  #  # #  # # # # #  # # ##    #   #  #     # #  #   ##  #  #     # #    #   # #  #  #   
%   ###  ##  #  #  # ###  # # # #  #  # #     ##  ##      #  ## #  ##   ## # ###   ### #   # ###    ## 
%                    #                  #                 #                                  #         
%                    #                ##                ##                                   #         
%
% =================================================================================================

JavaScript has become the \emph{lingua franca} of the web and increasing number of programming 
languages provide a way of compiling to JavaScript. In F\#, the first project was F\# WebTools
\cite{fsharp-webtools} in 2007. A more recent and complex framework called Websharper \cite{websharper-guis} 
is available with full commercial support. In this paper, we use FunScript which is a lightweight
library focused just on translating F\# to JavaScript.

In this section, we discuss the options available when targeting JavaScript. The compilation to 
JavaScript can be implemented as compiler back-end or as a library based on meta-programming 
(Section~\ref{sec:js-meta}), there are different approaches to base libraries (Section~\ref{sec:js-lib}),
asynchronous computing (Section~\ref{sec:js-async}) and semantic mismatch between the host language
and JavaScript (Section~\ref{sec:js-relative}).

\subsection{Lightweight meta-programming with quotations}
\label{sec:js-meta}

The FunScript library is based on F\# quotations \cite{fsharp-metaprog}. This means that the code
is compiled as ordinary F\#, but the compiler also stores marked blocks of code as data. In FunScript,
we use a launcher that, when the program runs, analyses its quotations and produces a JavaScript file
(so, the workflow is to compile the code as usual and then run it to produce JavaScript).

In FunScript, we typically need to translate the entire source file. To allow this, we instruct the
compiler to store code as data for the whole module using the \ident{ReflectedDefinition} attribute.
This is done at the beginning of the file (which we omitted in Section~\ref{sec:case}):
%
\begin{equation*}
\begin{array}{l}
 \lbrack\langl\ident{ReflectedDefinition}\rangl\rbrack \\
 \kvd{module}~\ident{Program}
 \\[0.5em]
 \kvd{open}~\ident{FunScript} \\
 \kvd{open}~\ident{FunScript.TypeScript} \\
 \kvd{open}~\ident{FSharp.Data} 
\end{array}
\end{equation*}
%
When marked with \ident{ReflectedDefinition}, the F\# compiler stores the body of all functions and 
methods in the marked module as \emph{quotations} that can be retrieved at run-time. Note that this does
not store the AST for the entire file -- for example, information about declared types need to be 
obtained separately using the .NET reflection mechanism.

The quotation mechanism has other uses. In the following, we use an explicit quotation $\langl @ \ldots @\rangl$,
which returns the wrapped code as data. This can be used, for example, to compile a matrix calculation
to run on a GPU \cite{accelerator}:
%
\begin{equation*}
\begin{array}{l}
 \langl @~ \kvd{fun}~(\ident{input}:\ident{Matrix}\langl\ident{float32}\rangl)~\rightarrow \\
 \qquad~~ \kvd{let}~\ident{sum}~=~(\ident{shift input}~-\hspace{-0.2em}1~0) ~+~ \ident{input} ~+~ (\ident{shift input}~+\hspace{-0.2em}1~0) \\
 \qquad~~ \ident{sum}~/~3.0\textnormal{f}~~@\rangl
\end{array}
\end{equation*}
%
Compared to writing a full compiler back-end, the lightweight approach to meta-programming makes it easy to 
write a translator from F\# to other languages. This has been used for compiling to SQL queries, compiling
F\# code to CUDA, for Freebase queries \cite{fsharp-data} and other applications.

Using the lightweight meta-programming approach in FunScript would not be appropriate if we wanted to 
compile arbitrary existing F\# source code to JavaScript without any modifications. For that, using a 
compiler back-end is a better choice. However, for the task solved in our case study (compiling 
newly written F\# code), the approach works well. A related question is providing mapping for 
standard (in our case, F\# and .NET) libraries used by the code. We discuss this in the next section.

\subsection{Accessing F\# and JavaScript libraries}
\label{sec:js-lib}

The F\# code in our case study uses a number of libraries. This includes the F\# core library (for 
example, the \ident{List.map} function), standard .NET libraries (iteration using the \kvd{for} loop
uses .NET \ident{IEnumerable} interface) and JavaScript libraries. Accessing these libraries in 
FunScript follows similar approach to F\# WebTools \cite{fsharp-webtools}.

The range of possible approaches to libraries when translating to JavaScript has two extreme cases.
We can attempt to port all libraries of our source language (F\#, Haskell or even .NET), or we 
can ignore most standard libraries and instead provide access to as many JavaScript libraries as 
possible. FunScript stands in the middle -- it provides access to \emph{some} F\# and .NET libraries,
but most advanced functionality is accessed through JavaScript libraries.

\paragraph{Mapping standard .NET and F\# libraries.}
The F\# ecosystem relies on .NET libraries. This means that taking an F\# library and translating it 
to JavaScript without any modification is not a viable approach as we would have to be able to also
translate any (compiled) .NET library. This is where F\# differs, for example, from Haskell which has
a closed ecosystem and the ghcjs project \cite{haskell-ghcjs} is thus capable of translating many 
standard libraries.

FunScript implements (explicit) mappings for standard F\# and .NET libraries that are useful in 
JavaScript (collections, string manipulation, date and other). Those are either reimplemented in
F\# with \ident{ReflectedDefinition} attribute, or mapped to standard JavaScript library.

An interesting aspect of the case study is that we use F\# Data \cite{fsharp-data} type providers, 
which is a standard F\# library that has not been built specifically for FunScript. This illustrates 
how type providers and quotations interact.
